---
title: "Project 7: Difference-in-Differences and Synthetic Control"
output: html_document
---


```{r}
# Install and load packages 
if (!require("pacman")) install.packages("pacman")

devtools::install_github("ebenmichael/augsynth")

pacman::p_load(# Tidyverse packages including dplyr and ggplot2 
               tidyverse,
               ggthemes,
               augsynth,
               gsynth, 
               directlabels)


# set seed
set.seed(44)

# load data
medicaid_expansion <- read_csv('medicaid_expansion.csv')
```


# Introduction

For this project, you will explore the question of whether the Affordable Care Act increased health insurance coverage (or conversely, decreased the number of people who are uninsured). The ACA was passed in March 2010, but several of its provisions were phased in over a few years. The ACA instituted the "individual mandate" which required that all Americans must carry health insurance, or else suffer a tax penalty. There are four mechanisms for how the ACA aims to reduce the uninsured population:

- Require companies with more than 50 employees to provide health insurance.
- Build state-run healthcare markets ("exchanges") for individuals to purchase health insurance.
- Provide subsidies to middle income individuals and families who do not qualify for employer based coverage.
- Expand Medicaid to require that states grant eligibility to all citizens and legal residents earning up to 138\% of the federal poverty line. The federal government would initially pay 100\% of the costs of this expansion, and over a period of 5 years the burden would shift so the federal government would pay 90\% and the states would pay 10\%.

In 2012, the Supreme Court heard the landmark case NFIB v. Sebelius, which principally challenged the constitutionality of the law under the theory that Congress could not institute an individual mandate. The Supreme Court ultimately upheld the individual mandate under Congress's taxation power, but struck down the requirement that states must expand Medicaid as impermissible subordination of the states to the federal government. Subsequently, several states refused to expand Medicaid when the program began on January 1, 2014. This refusal created the "Medicaid coverage gap" where there are indivudals who earn too much to qualify for Medicaid under the old standards, but too little to qualify for the ACA subsidies targeted at middle-income individuals.

States that refused to expand Medicaid principally cited the cost as the primary factor. Critics pointed out however, that the decision not to expand primarily broke down along partisan lines. In the years since the initial expansion, several states have opted into the program, either because of a change in the governing party, or because voters directly approved expansion via a ballot initiative.

You will explore the question of whether Medicaid expansion reduced the uninsured population in the U.S. in the 7 years since it went into effect. To address this question, you will use difference-in-differences estimation, and synthetic control.

# Data

The dataset you will work with has been assembled from a few different sources about Medicaid. The key variables are:

- **State**: Full name of state
- **Medicaid Expansion Adoption**: Date that the state adopted the Medicaid expansion, if it did so.
- **Year**: Year of observation.
- **Uninsured rate**: State uninsured rate in that year.

# Exploratory Data Analysis

Create plots and provide 1-2 sentence analyses to answer the following questions:

- Which states had the highest uninsured rates prior to 2014? The lowest?
- Which states were home to most uninsured Americans prior to 2014? How about in the last year in the data set? **Note**: 2010 state population is provided as a variable to answer this question. In an actual study you would likely use population estimates over time, but to simplify you can assume these numbers stay about the same.

```{r}
#explore data
head(medicaid_expansion)
summary(medicaid_expansion)
```


```{r}
# highest and lowest uninsured rates prior to 2014

#bar plots for 2008-2013

#2008
bar_2008 <- ggplot(medicaid_expansion %>% filter(year == '2008'),
                   aes(State, uninsured_rate)) + geom_col() +
  ggtitle('Uninsured Rate Per State in 2008') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2008

#2009
bar_2009 <- ggplot(medicaid_expansion %>% filter(year == '2009'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2009') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2009

#2010
bar_2010 <- ggplot(medicaid_expansion %>% filter(year == '2010'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2010') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2010

#2011
bar_2011 <- ggplot(medicaid_expansion %>% filter(year == '2011'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2011') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2011

#2012
bar_2012 <- ggplot(medicaid_expansion %>% filter(year == '2012'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2012') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2012

#2013
bar_2013 <- ggplot(medicaid_expansion %>% filter(year == '2013'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2013') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2013


```
Before 2014, the state that had the highest uninsured rate varied slightly. From 2008 to 2009, Utah had the highest uninsured rate, but from 2010 to 2013, Texas had the highest uninsured rate. From 2008 to 2013, Nevada had the second-highest uninsured rate. Alternatively, Massachusetts consistently had the lowest uninsured rate across all states for the entire period reported before 2014 (2008-2013). 


```{r}
# most uninsured Americans

#create new variable for uninsured Americans per state
medicaid_expansion$pop_uninsured <- medicaid_expansion$uninsured_rate * medicaid_expansion$population 

minmax <- medicaid_expansion %>% group_by(year) %>% 
  summarize(min = min(pop_uninsured), 
            minState = State[which.min(pop_uninsured)],
            max =max(pop_uninsured), 
            maxState = State[which.max(pop_uninsured)])
View(minmax)
```

Unsurprisingly, California, the state with the largest population in the US, had the largest uninsured population from 2008 to 2013. However, starting 2014, with California's adoption of the Medicaid Expansion, Texas had the most uninsured population until 2020 (the last year of available data). On the other hand, Wyoming had the least uninsured population from 2008 to 2009, but that shifted to Vermont from 2010 until 2020. Wyoming and Vermont are both states with small populations.

# Difference-in-Differences Estimation

## Estimate Model

Do the following:

- Choose a state that adopted the Medicaid expansion on January 1, 2014 and a state that did not. **Hint**: Do not pick Massachusetts as it passed a universal healthcare law in 2006, and also avoid picking a state that adopted the Medicaid expansion between 2014 and 2015.

-- States that adopted the expansion on January 1, 2014 include Arizona, Arkansas, California, Delaware, DC, Hawaii, Illinois, Iowa, Maryland, Minnesota, Nevada, New Jersey, New Mexico, New York, North Dakota, Ohio, Oregon, Rhode Island, Vermont, Washington, West Virginia. (Do not pick Kentucky because the date of adoption is listed at January 1, 2014, but in 2015 Kentucky got a new governor who dismantled it.)
-- States that adopted the expansion later: Alaska, Indiana, Louisiana, Michigan, Montana, Nebraska, New Hampshire, Pennsylvania, Utah, Virginia
-- States that never adopted the expansion: Alabama, Florida, Georgia, Kansas, Maine, Mississippi, Missouri, North Carolina, Oklahoma, South Carolina, South Dakota, Tennessee, Texas, Wisconsin, Wyoming.

- Assess the parallel trends assumption for your choices using a plot. If you are not satisfied that the assumption has been met, pick another state and try again (but detail the states you tried).

```{r}
# Parallel Trends plot

#We also tried California-Texas, California-Pennsylvania

medicaid_expansion %>%
  filter(State %in% c("California","Florida")) %>%
  ggplot() + 
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = State)) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = State)) +
  geom_vline(aes(xintercept = 2014)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('CA & FL Uninsured Rate Before and After Medicaid Expansion in 2014') +
  xlab('Year') +
  ylab('Uninsured Rate')
```
The parallel trends plot for the uninsured rate over time between California and Florida seems to meet the assumption that the trends between the two states were relatively parallel prior to the treatment year, showing fairly similar slopes between 2008 to 2014. 

- Estimates a difference-in-differences estimate of the effect of the Medicaid expansion on the uninsured share of the population. You may follow the lab example where we estimate the differences in one pre-treatment and one post-treatment period, or take an average of the pre-treatment and post-treatment outcomes

```{r}
# Difference-in-Differences estimation for uninsured rate

# California - Florida 
states_DiD <- medicaid_expansion %>%
  filter(State %in% c("California","Florida")) %>%
  filter(year >= 2013 & year<= 2015) 

#pre-treatment difference, uninsured rate
pre_diff_rate <- states_DiD %>%
  filter(year == 2013) %>%
  select(State, 
         uninsured_rate) %>%
  spread(State, 
         uninsured_rate) %>%
  summarise(California - Florida) 
pre_diff_rate
  
# post-treatment difference, uninsured rate

post_diff_rate <- states_DiD %>%
  filter(year == 2015) %>%
  select(State, 
         uninsured_rate) %>%
  spread(State, 
         uninsured_rate) %>%
  summarise(California - Florida) 
post_diff_rate

# diff-in-diffs, uninsured rate

diff_in_diffs_rate <- post_diff_rate - pre_diff_rate
diff_in_diffs_rate

```
Looking at the difference-in-differences estimation for the uninsured rate, the adoption of 2014 Medicaid expansion in California had a treatment effect of -0.02, or a 2% decrease in the rate of uninsured Americans.

## Discussion Questions

- Card/Krueger's original piece utilized the fact that towns on either side of the Delaware river are likely to be quite similar to one another in terms of demographics, economics, etc. Why is that intuition harder to replicate with this data?
- **Answer**:
The intuition that neighboring towns are similar to each other is harder to replicate with this data because the units are states, which are vastly different in demographics, economy, etc. More significantly, states have their own governing bodies determining the adoption of federal legislation like the Affordable Care Act that are unrelated to their geographic location. 

- What are the strengths and weaknesses of using the parallel trends assumption in difference-in-differences estimates?

- **Answer**:
A key strength of using the parallel trends assumption in DiD estimates is that we can use the untreated or control group as the counterfactual to the treated group and visually ascertain that trends over time between groups before treatment are generally constant. If the assumption is met, we can take the pre and post-treatment of the untreated group and compare it to the treated group to measure the treatment effect. The assumption also considers both time and selection bias and produces one estimate.

A critical weakness of this assumption is that there is no method to test its validity statistically. It also ignores possible confounding differences between the treatment and control groups. Furthermore, how we define the length of time may bias the estimates since the treatment effect varies depending on the period of time observed. 

# Synthetic Control

Estimate Synthetic Control

Although several states did not expand Medicaid on January 1, 2014, many did later on. In some cases, a Democratic governor was elected and pushed for a state budget that included the Medicaid expansion, whereas in others voters approved expansion via a ballot initiative. The 2018 election was a watershed moment where several Republican-leaning states elected Democratic governors and approved Medicaid expansion. In cases with a ballot initiative, the state legislature and governor still must implement the results via legislation. For instance, Idaho voters approved a Medicaid expansion in the 2018 election, but it was not implemented in the state budget until late 2019, with enrollment beginning in 2020.

Do the following:

- Choose a state that adopted the Medicaid expansion after January 1, 2014. Construct a non-augmented synthetic control and plot the results (both pre-treatment fit and post-treatment differences). Also report the average ATT and L2 imbalance.

```{r}
# non-augmented synthetic control

# Chosen state: Louisiana

# Add a pre/post-treatment indicator
medicaid_expansion$Year_Adopted = format(medicaid_expansion$Date_Adopted, format="%Y")
medicaid_expansion$Year_Adopted = as.numeric(medicaid_expansion$Year_Adopted)
medicaid_expansion$treatment = if_else(medicaid_expansion$Year_Adopted > medicaid_expansion$year, 0, 1, missing=0)

# Add indicator for *LOUISIANA* PRE AND POST TREATMENT
medicaid_expansion$treatLou = ifelse(medicaid_expansion$State == "Louisiana" & medicaid_expansion$year > 2016, 1, 0)

# Create a synthetic control for Louisiana
syn_Lou = gsynth(uninsured_rate ~ treatLou, data=medicaid_expansion, Y=uninsured_rate, D=treatLou, index=c("State", "year"), seed = 5, min.T0 = 2, CV = FALSE)
syn_Lou

syn_Lou$att.avg # ATT = -0.07145
syn_Lou$sigma2 # Mean squared error of interactive fixed effect model = 0.0009858552
```
```{r}
# Plot the results (both pre-treatment fit and post-treatment differences).

plot(syn_Lou)

sum_syn_Lou <- summary(syn_Lou)

medicaid_syn_Lou <- medicaid_expansion %>%
  filter(State == "Louisiana") %>%
  bind_cols(difference = syn_Lou$att) %>%
  mutate(synthetic_Lou = uninsured_rate + difference)

# Plot

medicaid_syn_Lou %>%
  ggplot() +
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = 'Louisiana')) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = 'Louisiana')) +
  geom_point(aes(x = year, 
                 y = synthetic_Lou, 
                 color = 'Synthetic Louisiana')) +
  geom_line(aes(x = year, 
                y = synthetic_Lou, 
                color = 'Synthetic Louisiana')) +
  scale_color_manual(values = c('Louisiana' = 'red', 'Synthetic Louisiana' = 'blue')) +
  geom_vline(aes(xintercept = 2016.5)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Louisiana and Synthetic Louisiana') +
  xlab('Year') +
  ylab('Uninsured Rate')
# The fit is not great because there is actually a staggered adoption of the treatment that is not being taken into consideration in this model.
```

- Re-run the same analysis but this time use an augmentation (default choices are Ridge, Matrix Completion, and GSynth). Create the same plot and report the average ATT and L2 imbalance.

```{r}
# augmented synthetic control
augsyn <- augsynth(uninsured_rate ~ treatment, State, year, medicaid_expansion,
                    progfunc = "ridge", scm = T)
augsyn
plot(augsyn)
# This ran as multisynth
# ATT is -0.010
# Plot doesn't know which state it should be making a synthetic control from.

augsyn_Lou <- augsynth(uninsured_rate ~ treatLou, State, year, medicaid_expansion,
                    progfunc = "ridge", scm = T)
augsyn_Lou
# This ran as single_augsynth
# ATT is -0.026

sum_augsyn_Lou = summary(augsyn_Lou)
sum_augsyn_Lou$average_att$Estimate # ATT is -0.02570572
sum_augsyn_Lou$l2_imbalance #L2 Imbalance is 0.00513934

plot(augsyn_Lou) # The fit of this graph is much better than when using gsynth.
```
```{r}
sum_augsyn_Lou <- summary(augsyn_Lou)

medicaid_augsyn_Lou <- medicaid_syn_Lou %>%
  filter(State == "Louisiana") %>%
  bind_cols(aug_difference = sum_augsyn_Lou$att$Estimate) %>%
  mutate(aug_synthetic_Lou = uninsured_rate + aug_difference)

medicaid_augsyn_Lou %>%
  ggplot() +
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = 'Louisiana')) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = 'Louisiana')) +
  geom_point(aes(x = year, 
                 y = synthetic_Lou, 
                 color = 'Synthetic Louisiana')) +
  geom_line(aes(x = year, 
                y = synthetic_Lou, 
                color = 'Synthetic Louisiana')) +
  geom_point(aes(x = year, 
                 y = aug_synthetic_Lou, 
                 color = 'Augmented Synthetic Louisiana')) +
  geom_line(aes(x = year, 
                y = aug_synthetic_Lou, 
                color = 'Augmented Synthetic Louisiana')) +
  scale_color_manual(values = c('Louisiana' = 'red', 'Synthetic Louisiana' = 'blue', 'Augmented Synthetic Louisiana' = 'green')) +
  geom_vline(aes(xintercept = 2016.5)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Louisiana, Synthetic Louisiana, and Augmented Synthetic Louisiana') +
  xlab('Year') +
  ylab('Uninsured Rate')

#Augmented Synthetic Louisiana is a much better pre-treatment match than the Synthetic.
```

- Plot barplots to visualize the weights of the donors.

```{r}
# barplots of weights
data.frame(augsyn_Lou$weights) %>%
  # change index to a column
  tibble::rownames_to_column('State') %>%
  ggplot() +
  # stat = identity to take the literal value instead of a count for geom_bar()
  geom_bar(aes(x = State, 
               y = augsyn_Lou.weights),
           stat = 'identity') +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        axis.text.x = element_text(angle = 90)) +
  ggtitle('Synthetic Control Weights') +
  xlab('State') +
  ylab('Weight') 
```

**HINT**: Is there any preprocessing you need to do before you allow the program to automatically find weights for donor states?

## Discussion Questions


- What are the advantages and disadvantages of synthetic control compared to difference-in-differences estimators?
- **Answer**: The synthetic control can meet the parallel assumptions trend more precisely.

- One of the benefits of synthetic control is that the weights are bounded between [0,1] and the weights must sum to 1. Augmentation might relax this assumption by allowing for negative weights. Does this create an interpretation problem, and how should we balance this consideration against the improvements augmentation offers in terms of imbalance in the pre-treatment period?
- **Answer**: It does create an interpretation problem if the graph/visual representation of weights does not display negative weights. Ultimately, I think that as long as the error terms are small and there is a match pre-treatment, it is not a problem. 

# Staggered Adoption Synthetic Control

## Estimate Multisynth

Do the following:

- Estimate a multisynth model that treats each state individually. Choose a fraction of states that you can fit on a plot and examine their treatment effects.

```{r}
# multisynth model states
```

- Estimate a multisynth model using time cohorts. For the purpose of this exercise, you can simplify the treatment time so that states that adopted Medicaid expansion within the same year (i.e. all states that adopted epxansion in 2016) count for the same cohort. Plot the treatment effects for these time cohorts.

```{r}
# multisynth model time cohorts
```

## Discussion Questions

- One feature of Medicaid is that it is jointly administered by the federal government and the states, and states have some flexibility in how they implement Medicaid. For example, during the Trump administration, several states applied for waivers where they could add work requirements to the eligibility standards (i.e. an individual needed to work for 80 hours/month to qualify for Medicaid). Given these differences, do you see evidence for the idea that different states had different treatment effect sizes?
- **Answer**: 

- Do you see evidence for the idea that early adopters of Medicaid expansion enjoyed a larger decrease in the uninsured population?
- **Answer**: 

# General Discussion Questions

- Why are DiD and  synthetic control estimates well suited to studies of aggregated units like cities, states, countries, etc?
- **Answer**:

- What role does selection into treatment play in DiD/synthetic control versus regression discontinuity? When would we want to use either method?
- **Answer**: